---
title: "cross-tissue DE analysis with fastTopics"
output: html_document
date: '2024-04-29'
author: 'Jing Gu'
---
# GoM DE analysis on u19 dataset

## Model fitting 

Parameters: 

N_updates = 150
N_topics = 12
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(Matrix)
library(fastTopics)
library(ggplot2)
library(cowplot)
library(poolr)
library(dplyr)
library(tidyr)
library(colorRamp2)
library(ComplexHeatmap)
set.seed(10)

# load data
counts<-readRDS("output/fastTopics/aggregated_lymph_scRNA_counts.RDS")
fit<-readRDS("output/fastTopics/fit_model_updates12_k150.RDS")
metadata<-readRDS("output/fastTopics/u19_cell_metadata.RDS")

# add sample covariates information
metadata$donor_id<-metadata$orig.ident
metadata$donor_id[metadata$full.ident=="spleens_2"]<-'3'
metadata$donor_id[metadata$full.ident=="spleens_3"]<-'6'
batch_ids<-c("batch1", "batch1", "batch2", "batch2", "batch3", "batch3", "batch1", "batch2", "batch3")
metadata$batch<-factor(metadata$full.ident, labels = batch_ids)


```


## Model evaluation
```{r}
fit<-readRDS("output/fastTopics/fit_model_updates12_k150.RDS")
```

check the convergence
```{r}
# check for convergence of model fitting 
plot_progress(fit,x = "iter",add.point.every = 10,colors = "black") +
  theme_cowplot(font_size = 10)

summary(fit)
```

```{r}
loglik <- loglik_multinom_topic_model(counts,fit)
pdat <- data.frame(loglik)
ggplot(pdat,aes(loglik)) +
  geom_histogram(bins = 64,color = "white",fill = "black", linewidth = 0.25) +
  labs(y = "number of cells") +
  theme_cowplot(font_size = 10)

```

```{r}
pdat <- data.frame(loglik = loglik,subpop = metadata$majority_voting)
celltypes<-setdiff(unique(pdat$subpop),
                   c("ILC3", "Epithelial cells", "Plasma cells",
                     "Classical monocytes"))

ggplot(pdat %>% filter(subpop %in% celltypes),aes(x = loglik,fill = subpop)) +
  geom_histogram(bins = 64,color = "white",size = 0.25) +
  labs(y = "number of cells") +
  theme_cowplot(font_size = 10)
```

## Visualize topics with structural plots

* plot by cell-types

```{r message=F}
clusters<-factor(metadata$majority_voting)
structure_plot(fit,topics = 1:12,gap = 25,
               grouping = clusters,
               verbose=FALSE) 
```

* plot by tissue

```{r message=F}
structure_plot(fit,topics = 1:12,gap = 25,
               grouping = metadata$tissue.ident,
               verbose=FALSE) 
```

* plot by tissue and cell-type

```{r message=F}
metadata$major<-factor(metadata$majority_voting, 
                       labels = c("NK", "NK", "other", "other",
                                  "other", "Memory_B", "Naive_B",
                                  "other", "Treg", "CD4_T",
                                  "CD8_T", "Th17"))
tissue_samples<-paste(metadata$tissue.ident, metadata$major, sep="_")
ordered_tissue_samples<-unlist(
  lapply(unique(metadata$major), function(i){c(paste0("lungs_",  i),
                                               paste0("spleens_", i)
                                               )}))
tissue_samples<-factor(tissue_samples,
                       levels = ordered_tissue_samples)
structure_plot(fit,topics = 1:12,gap = 25,
               grouping = tissue_samples,
               verbose=FALSE
               ) 
```

## Validate topics with enrichment test

### GO enrichemnt test on top 500 genes ranked by loadings
```{r}
# set.seed(10)
# de <- de_analysis(fit,counts,pseudocount = 0.1,
#                   control = list(ns = 1e4,nc = 20))
# saveRDS(de, "output/fastTopics/GoM_DE_u19.RDS")
db <- c(#"pathway_KEGG", "disease_GLAD4U", "disease_OMIM",
                "Biological_Process", 
                "Cellular_Component",
                "Molecular_Function")

enrich_out<-readRDS("output/fastTopics/GO_enrichment_results_top500.RDS")
enrich_out<-data.frame(enrich_out, row.names = NULL)
DT::datatable(
  enrich_out %>% 
    filter(database==sprintf("geneontology_%s_noRedundant", db[3])) %>%
    mutate(FDR=format(FDR, scientific=TRUE, digits=2),
           enrichment=format(enrichmentRatio, digits=2)) %>%
    select(c(k, description, enrichment, FDR)),
   caption=db[3]
  )
```

```{r}
DT::datatable(
  enrich_out %>% 
    filter(database==sprintf("geneontology_%s_noRedundant", db[1])) %>%
    mutate(FDR=format(FDR, scientific=TRUE, digits=2),
           enrichment=format(enrichmentRatio, digits=2)) %>%
    select(c(k, description, enrichment, FDR)),
   caption=db[1]
  )
```

### GO enrichemnt test on GoM DE analysis

**Volcano plots for GoM DE results**

The z-scores for posterior mean log-fold change estimates vs. log-fold change
```{r}
de<-readRDS("output/fastTopics/GoM_DE_u19.RDS")
de_list<-list()
for(k in 1:12){
  dat <- data.frame(postmean = de$postmean[,k],
                  z        = de$z[,k],
                  lfsr     = de$lfsr[,k])
  
  dat <- subset(dat,lfsr < 0.01)
  dat <- dat[order(dat$postmean,decreasing = TRUE),]
  de_list[[k]]<-dat
  print(volcano_plot(de, k=k, ymax=250,
  plot.title = sprintf("topic:%s", k)))
}
```
**GO Enrichemnt results for upregulated genes with local false sign rates < 0.01**

Topic 3,4,5,6 show strong enrichment for the GO terms of T-cell activation, leukocyte differentiation, positive regulation of cell adhesion, adaptive immune response. 

From volcano plots, we see topic 6 has high LFC in genes such as IL17A, IL22, which are mainly produced by Th17 cells. 

Topic 11 is strongly enriched for GO terms related to regulation of innate or adaptive immune response. 

Topic 12 shows strong enrichment for the GO term of B cell activation.

```{r}
enrich_out<-readRDS("output/fastTopics/GO_enrichment_results_GoM_DE.RDS")
enrich_out<-data.frame(enrich_out, row.names = NULL)
DT::datatable(
  enrich_out %>% 
    filter(database==sprintf("geneontology_%s_noRedundant", db[3])) %>%
    mutate(FDR=format(FDR, scientific=TRUE, digits=2),
           enrichment=format(enrichmentRatio, digits=2)) %>%
    select(c(k, description, enrichment, FDR)),
   caption=db[3]
  )
```

```{r}
DT::datatable(
  enrich_out %>% 
    filter(database==sprintf("geneontology_%s_noRedundant", db[1])) %>%
    mutate(FDR=format(FDR, scientific=TRUE, digits=2),
           enrichment=format(enrichmentRatio, digits=2)) %>%
    select(c(k, description, enrichment, FDR)),
   caption=db[1]
  )
```


```{r}
DT::datatable(
  enrich_out %>% 
    filter(database==sprintf("geneontology_%s_noRedundant", db[2])) %>%
    mutate(FDR=format(FDR, scientific=TRUE, digits=2),
           enrichment=format(enrichmentRatio, digits=2)) %>%
    select(c(k, description, enrichment, FDR)),
   caption=db[2]
  )
```

## Identify topics correlated with differences between tissue

1. test whether any topic is associated with transcriptional differences across tissue
$$
L  = \beta X_{\text{tissue}} + \text{Covariates} + \epsilon
$$
2. perform T-test to see whether topic proportions between two tissues are significantly different

### Density plot for topic proportions between tissue across cell types

X-axis denotes the percentage format of topic proportion.  
Y-axis represents the density of cells with respect to topic proportions.

Topic 1 is specific to cell types other than lymphocytes.

Topic 3 is highly represented in NK and $CD8^+$ T cells.

Topic 4-5 are T-cell specific, with clear separation between tissue in $CD8^+$ T for topic 4 and Treg for topic 5.

Topic 6 is specific to Th17 cells and lungs has a longer tail of distribution compared to spleens.

Topic 2,7,8,9 show similar distributions across cell types without clear separation between tissue. 

Topic 10-12 show higher density in B cells. Topic 11 captures tissue difference mainly in Memory B cells, while topic 12 displays some separation between tissue in Naive B cells.

```{r}
# convert poisson NMF model to multinational model
multinom_fit<-poisson2multinom(fit)
pdat<-data.frame(multinom_fit$L)
pdat$tissue<-metadata$tissue.ident
pdat$celltype<-metadata$major
pdat<-pdat %>% 
  pivot_longer(
    !c(tissue, celltype),
    names_to="topic", values_to="proportion")

# compare by tissue
#plt_list <- list()
for (k in unique(pdat$topic)) {
  print(
    ggplot(pdat %>% filter(topic == k),
           aes(proportion * 100, fill = tissue)) +
      geom_density(alpha = 0.4) +
      facet_grid(vars(celltype), scales = "free_y") +
      theme(strip.text =
              element_text(size = 8)) +
      xlab("percentage") +
      ggtitle(paste0("Topic:", k))
  )
}

```

### Barplot for topic proportions in cell types

* The topic proportion for k10 is higher in spleens than lungs in both naive B and memory B cells.
* The topic proportions for k11 is higher in spleens than lungs in memory B cells only.
* The topic proportions for k4 is higher in spleens than lungs across cell types.
* The topic proportions for k6 is higher in lungs than spleens across cell types.
* The topic proportions for k5 is higher in lungs than spleens across T cell subsets.

```{r}
# cell number by tissue and cell type
table(metadata$tissue.ident, metadata$major)

# Sum of topic proportion by tissue, cell type and topic
pdat %>% group_by(tissue, celltype, topic) %>% summarize(topic_prop_sum=sum(proportion))

# Scaled to be added up to be 1
prop<- pdat %>% group_by(tissue, celltype, topic) %>% summarize(topic_prop=sum(proportion)/n())

prop_sum<- pdat %>% group_by(tissue, celltype, topic) %>% summarize(topic_prop=sum(proportion))

ggplot(prop, aes(x = tissue, y = topic_prop, fill = topic)) +
  geom_bar(stat = "identity") + facet_grid(cols = vars(celltype)) + xlab("") + 
theme(axis.text.x = element_text(angle = 90)) + ylab("topic proportion")

```
All k4 terms included in k5.
```{r}
# check differences in GO terms between topics
# k4 vs k5
go_bp<-enrich_out %>% filter(database=="geneontology_Biological_Process_noRedundant") %>% select(-c(userId))
go_diff<-setdiff(go_bp %>% filter(k=="k5") %>% select(description),
                 go_bp %>% filter(k=="k4") %>% select(description))
go_bp %>% filter(description %in% go_diff$description & k=="k5") %>% select(k, description, FDR)
```

```{r}
# k6 vs.k5
go_diff<-setdiff(go_bp %>% filter(k=="k5") %>% select(description),
                 go_bp %>% filter(k=="k6") %>% select(description))
head(go_diff)

# k5 vs. k6
go_diff<-setdiff(go_bp %>% filter(k=="k6") %>% select(description),
                 go_bp %>% filter(k=="k5") %>% select(description))
go_bp %>% filter(description %in% go_diff$description & k=="k6") %>% select(k, description, FDR)

```

All k10 terms included in k11.
```{r}
#k11 vs. k10
go_diff<-setdiff(go_bp %>% filter(k=="k11") %>% select(description),
                 go_bp %>% filter(k=="k10") %>% select(description))
go_bp %>% filter(description %in% go_diff$description & k=="k11") %>% select(k, description, FDR)

```



## Perform t-test while adjusting for confounders


### Procedure

Test mean difference between tissue one donor at a time and then do meta-analysis with Fisher's method

### Results
X-axis denotes cell types and y-axis denotes the topics. For major cell types, we saw majority of topics have significant differences in proportions between tissue. 

```{r}
pdat<-data.frame(multinom_fit$L)
pdat$tissue<-metadata$tissue.ident
pdat$celltype<-metadata$major
pdat$donor_id<-metadata$donor_id
pdat<-pdat %>% 
  pivot_longer(
    !c(tissue, celltype, donor_id),
    names_to="topic", values_to="proportion")

celltypes<-setdiff(unique(pdat$celltype),
                    c("Treg", "Th17")
                    )
test_meandiff_by_celltype<-function(pdat){
  out_list<-list()
  for (j in setdiff(unique(pdat$celltype),
                    c("Treg", "Th17")
                    )){
    pdat_sub<-pdat %>% filter(celltype == j)
    out<-c()
    for(k in unique(pdat_sub$topic)){
      model<-t.test(pdat_sub %>% filter(tissue =="lungs" & topic == k) %>% 
                      select(proportion),
                    pdat_sub %>% filter(tissue =="spleens" & topic == k) %>% 
                      select(proportion))
      out<-rbind(out, c(k, model$estimate, model$p.value))
    }
    out<-data.frame(out)
    colnames(out)<-c("topic", "mean_lungs", "mean_spleens", "pval")
    out_list[[j]]<-out
  }
  return(out_list)
}
  
ranksum_test_by_celltype<-function(pdat){
  out_list<-list()
  for (j in setdiff(unique(pdat$celltype),
                    c("Treg", "Th17")
                    )){
    pdat_sub<-pdat %>% filter(celltype == j)
    out<-c()
    for(k in unique(pdat_sub$topic)){
      model <-
        wilcox.test(
          (pdat_sub %>% filter(tissue == "lungs" &
                                topic == k) %>% select(proportion))$proportion,
          (pdat_sub %>% filter(tissue == "spleens" &
                                topic == k) %>% select(proportion))$proportion
          
        )
      out<-rbind(out, c(k, model$p.value))
    }
    out<-data.frame(out)
    colnames(out)<-c("topic", "pval")
    out_list[[j]]<-out
  }
  return(out_list)
}

donor1<-test_meandiff_by_celltype(pdat%>%filter(donor_id=="1"))
donor2<-test_meandiff_by_celltype(pdat%>%filter(donor_id=="3"))
donor3<-test_meandiff_by_celltype(pdat%>%filter(donor_id=="6"))


combined_pvals<-list()
for(i in celltypes){
  combined_pvals[[i]]<-c()
  for(j in 1:length(unique(pdat$topic))){
    model<-fisher(
      as.numeric(
          donor1[[i]][j, "pval"],
          donor2[[i]][j, "pval"],
          donor3[[i]][j, "pval"])
      )
    combined_pvals[[i]]<-c(combined_pvals[[i]], model$p)
  }
}

```

```{r}
df<-do.call(rbind, combined_pvals)
colnames(df)<-paste0("k", 1:dim(df)[2])
col_fun_prop = colorRamp2(c(0, 0.05, 1), c("blue", "white", "red"))
plt <- Heatmap(df, col=col_fun_prop, 
               cluster_rows=FALSE,
               cluster_columns=FALSE,
               heatmap_legend_param = list(
                 col_fun=col_fun_prop, 
                 title="combined p-value",
                 at = c(0, 0.05, 1)
               ))
plt
```






