---
title: "cross-tissue DE analysis with fastTopics"
output: html_document
date: '2024-04-29'
author: 'Jing Gu'
---
# GoM DE analysis on u19 dataset

## Model fitting 

Parameters: 

N_updates = 150
N_topics = 12
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(Matrix)
library(fastTopics)
library(ggplot2)
library(cowplot)
library(poolr)
library(dplyr)
library(tidyr)
library(colorRamp2)
library(ComplexHeatmap)
set.seed(10)

# load data
counts<-readRDS("output/fastTopics/aggregated_lymph_scRNA_counts.RDS")
fit<-readRDS("output/fastTopics/fit_model_updates12_k150.RDS")
metadata<-readRDS("output/fastTopics/u19_cell_metadata.RDS")

# add sample covariates information
metadata$donor_id<-metadata$orig.ident
metadata$donor_id[metadata$full.ident=="spleens_2"]<-'3'
metadata$donor_id[metadata$full.ident=="spleens_3"]<-'6'
batch_ids<-c("batch1", "batch1", "batch2", "batch2", "batch3", "batch3", "batch1", "batch2", "batch3")
metadata$batch<-factor(metadata$full.ident, labels = batch_ids)


```


## Model evaluation
```{r}
fit<-readRDS("output/fastTopics/fit_model_updates12_k150.RDS")
```

check the convergence
```{r}
# check for convergence of model fitting 
plot_progress(fit,x = "iter",add.point.every = 10,colors = "black") +
  theme_cowplot(font_size = 10)

summary(fit)
```

```{r}
loglik <- loglik_multinom_topic_model(counts,fit)
pdat <- data.frame(loglik)
ggplot(pdat,aes(loglik)) +
  geom_histogram(bins = 64,color = "white",fill = "black", linewidth = 0.25) +
  labs(y = "number of cells") +
  theme_cowplot(font_size = 10)

```

```{r}
pdat <- data.frame(loglik = loglik,subpop = metadata$majority_voting)
celltypes<-setdiff(unique(pdat$subpop),
                   c("ILC3", "Epithelial cells", "Plasma cells",
                     "Classical monocytes"))

ggplot(pdat %>% filter(subpop %in% celltypes),aes(x = loglik,fill = subpop)) +
  geom_histogram(bins = 64,color = "white",size = 0.25) +
  labs(y = "number of cells") +
  theme_cowplot(font_size = 10)
```

## Visualize topics with structural plots

* plot by cell-types

```{r message=F}
clusters<-factor(metadata$majority_voting)
structure_plot(fit,topics = 1:12,gap = 25,
               grouping = clusters,
               verbose=FALSE) 
```

* plot by tissue

```{r message=F}
structure_plot(fit,topics = 1:12,gap = 25,
               grouping = metadata$tissue.ident,
               verbose=FALSE) 
```

* plot by tissue and cell-type

```{r message=F}
metadata$major<-factor(metadata$majority_voting, 
                       labels = c("NK", "NK", "other", "other",
                                  "other", "Memory_B", "Naive_B",
                                  "other", "Treg", "CD4_T",
                                  "CD8_T", "Th17"))
tissue_samples<-paste(metadata$tissue.ident, metadata$major, sep="_")
ordered_tissue_samples<-unlist(
  lapply(unique(metadata$major), function(i){c(paste0("lungs_",  i),
                                               paste0("spleens_", i)
                                               )}))
tissue_samples<-factor(tissue_samples,
                       levels = ordered_tissue_samples)
structure_plot(fit,topics = 1:12,gap = 25,
               grouping = tissue_samples,
               verbose=FALSE
               ) 
```

## Validate topics with enrichment test
```{r eval=F}
set.seed(10)
de <- de_analysis(fit,counts,pseudocount = 0.1,
                  control = list(ns = 1e4,nc = 20))
```



## Identify topics correlated with differences between tissue

1. test whether any topic is associated with transcriptional differences across tissue
$$
L  = \beta X_{\text{tissue}} + \text{Covariates} + \epsilon
$$
2. perform T-test to see whether topic proportions between two tissues are significantly different

### Density plot for topic proportions between tissue across cell types

X-axis denotes percentage format of topic proportion
```{r}
# convert poisson NMF model to multinational model
multinom_fit<-poisson2multinom(fit)
pdat<-data.frame(multinom_fit$L)
pdat$tissue<-metadata$tissue.ident
pdat$celltype<-metadata$major
pdat$donor_id<-metadata$donor_id
pdat<-pdat %>% 
  pivot_longer(
    !c(tissue, celltype, donor_id),
    names_to="topic", values_to="proportion")
# compare by tissue
#plt_list <- list()
for (k in unique(pdat$topic)) {
  print(k)
  print(
    ggplot(pdat %>% filter(topic == k),
         aes(proportion * 100, fill = tissue)) +
    geom_density(alpha = 0.4) +
    facet_grid(vars(celltype)) +
    theme(strip.text =
            element_text(size = 10)) +
    xlab("percentage")
  )
}

```

## Perform t-test while adjusting for confounders


### Procedure

Test mean difference between tissue one donor at a time and then do meta-analysis with Fisher's method

### Results
X-axis denotes cell types and y-axis denotes the topics. For major cell types, we saw majority of topics have significant differences in proportions between tissue. 

```{r}
celltypes<-setdiff(unique(pdat$celltype),
                    c("Treg", "Th17")
                    )
test_meandiff_by_celltype<-function(pdat){
  out_list<-list()
  for (j in setdiff(unique(pdat$celltype),
                    c("Treg", "Th17")
                    )){
    pdat_sub<-pdat %>% filter(celltype == j)
    out<-c()
    for(k in unique(pdat_sub$topic)){
      model<-t.test(pdat_sub %>% filter(tissue =="lungs" & topic == k) %>% 
                      select(proportion),
                    pdat_sub %>% filter(tissue =="spleens" & topic == k) %>% 
                      select(proportion))
      out<-rbind(out, c(k, model$estimate, model$p.value))
    }
    out<-data.frame(out)
    colnames(out)<-c("topic", "mean_lungs", "mean_spleens", "pval")
    out_list[[j]]<-out
  }
  return(out_list)
}
  
ranksum_test_by_celltype<-function(pdat){
  out_list<-list()
  for (j in setdiff(unique(pdat$celltype),
                    c("Treg", "Th17")
                    )){
    pdat_sub<-pdat %>% filter(celltype == j)
    out<-c()
    for(k in unique(pdat_sub$topic)){
      model <-
        wilcox.test(
          (pdat_sub %>% filter(tissue == "lungs" &
                                topic == k) %>% select(proportion))$proportion,
          (pdat_sub %>% filter(tissue == "spleens" &
                                topic == k) %>% select(proportion))$proportion
          
        )
      out<-rbind(out, c(k, model$p.value))
    }
    out<-data.frame(out)
    colnames(out)<-c("topic", "pval")
    out_list[[j]]<-out
  }
  return(out_list)
}

donor1<-test_meandiff_by_celltype(pdat%>%filter(donor_id=="1"))
donor2<-test_meandiff_by_celltype(pdat%>%filter(donor_id=="3"))
donor3<-test_meandiff_by_celltype(pdat%>%filter(donor_id=="6"))

combined_pvals<-list()
for(i in celltypes){
  combined_pvals[[i]]<-c()
  for(j in 1:length(unique(pdat$topic))){
    model<-fisher(
      as.numeric(
          donor1[[i]][j, "pval"],
          donor2[[i]][j, "pval"],
          donor3[[i]][j, "pval"])
      )
    combined_pvals[[i]]<-c(combined_pvals[[i]], model$p)
  }
}

```

```{r}
df<-do.call(rbind, combined_pvals)
colnames(df)<-paste0("k", 1:dim(df)[2])
col_fun_prop = colorRamp2(c(0, 0.05, 1), c("blue", "white", "red"))
plt <- Heatmap(df, col=col_fun_prop, 
               cluster_rows=FALSE,
               cluster_columns=FALSE,
               heatmap_legend_param = list(
                 col_fun=col_fun_prop, 
                 title="combined p-value",
                 at = c(0, 0.05, 1)
               ))
plt
```






