---
title: "Differential peak analyses"
output: html_document
date: '`r Sys.Date()`'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(ArchR)
library(dplyr)
library(tidyr)
library(rtracklayer)
library(GenomicRanges)
library(colorRamp2)
library(scales)
library(rGREAT)
library(edgeR)
library(ggrepel)
library(BSgenome.Hsapiens.UCSC.hg38)
library(ComplexHeatmap)
#library(plotly)
#source("~/projects/lung_immune_fine_mapping/code/archR_helper.r")
set.seed(1)
options(repr.plot.width=11,.plot.height=8)

addArchRGenome("hg38")
work.dir<-"~/projects/u19_multiomics/data/cellranger_outs"

#proj <- loadArchRProject("~/cluster/projects/u19_multiomics/analyses/u19_peaks_tissue/")
ct_order <-
  c(
  "Naive_B",
  "Memory_B",
  "NK",
  "CD8_T",
  "CD8/CD4_T",
  "CD4_T",
  "Th17/CD4_T",
  "Treg",
  "Other"
)
```

# Objective

* Identify genomic regions with differential accessibility across tissue or disease status
* Fine-mapped variants within cell-type specific peaks may suggest specific roles of immune subsets in asthma
* Fine-mapped variants within cross-tissue/status DA peaks in specific immune subsets may further provide the direction of the effects on genetic risks of asthma

## Cell composition by donor

Since COB-8 sample has nearly no lymphocytes, we decided to remove COB-8 from analyses. 
```{r}
metadata <- readRDS("output/u19_output/u19_full_atac_cell_metadata.RDS")
gginput <- 
  data.frame(metadata) %>%
  mutate(major = 
           factor(PredCellType,
                  labels = c("T", "T", "T",
                             "B", "NK", "B",
                             "Other", "T", "T")))
  
ggplot(
  gginput,
  aes( y = paste(Tissue, Sample, sep="_"),
      fill = major)) + ylab("") +
  geom_bar()

totCounts <- 
  data.frame(
    table(metadata$Sample)) %>%
    mutate(Sample = Var1,
           nCells = Freq)
totCounts$Tissue <- metadata$Tissue[match(totCounts$Sample,
                                          metadata$Sample)]
tbl <- 
  left_join(
    data.frame(table(metadata$PredCellType,
                   metadata$Sample)) %>%
    mutate(
      CellType = Var1,
      Sample = Var2,
      Counts = Freq),
    totCounts, by = "Sample")


ggplot(tbl %>%
         mutate(CellType =
                  factor(CellType,
                         labels = c("T", "T", "T",
                                     "B", "NK", "B",
                                    "Other", "T", "T"))),
       aes(
         y = paste(Tissue, Sample, sep = "_"),
         x = Counts / nCells,
         fill = CellType
       )) +
  xlab("Cell Proportion") + ylab("") +
  geom_col()
```

```{r eval = F }
# differential analyses with Libra package
atac_input <- readRDS("u19_full_atac_by_tissue-celltype_raw_counts.RDS")
metadata <- readRDS("u19_full_atac_cell_metadata.RDS")
rownames(atac_input) <- as.character(1:dim(atac_input)[1])
metadata$major <- 
  factor(metadata$PredCellType,
         labels = c("B", "B", "NK",
                    "T", "T", "T", "T", "T", "Other"))
metadata$batch <- ifelse(grepl("SMO", metadata$Sample), 1, 0)
cross_tissue <- 
  run_de(
    input = atac_input,
    meta = data.frame(metadata),
    replicate_col = "Sample",
    cell_type_col = "major",
    label_col = "Tissue",
    min_cells = 3,
    min_reps = 2,
    min_features = 2,
    de_family = "singlecell",
    de_method = "permutation",
   # de_type = "LRT",
    input_type = "scATAC",
    normalization = "log_tp10k",
    binarization = FALSE,
    latent_vars = "batch",
    n_threads = 6
  )
```

## Approach

**Models for EdgeR**

Modeling mean-variance relationship using a quasi-NB GLM
$$
\begin{eqnarray*} {\rm var}(y_{gi}) = \sigma ^2_g\mu _{gi} + \psi _g \mu _{gi}^2 \end{eqnarray*}

$$
**MDS plot**
distances between samples correspond to leading biological coefficient of variation (BCV) between those samples

```{r eval = F}
library(edgeR)
load("output/u19_output/u19_full_atac_by_tissue-celltype_pseudobulk.RData")
metadata$PredCellType <-
  factor(metadata$PredCellType,
         levels = ct_order)

# avoid over-correction of batch effects (correct for two batches)
metadata$batch <- ifelse(grepl("COB", metadata$Sample), 1, 0)
metadata$major <- 
  factor(metadata$PredCellType,
         labels = c("B", "B", "NK", "T", "T", "T", "T", "T", "Other"))

metadata$majorID <- 
  paste(metadata$Sample,
        metadata$major,
        sep = "_")
metadata$PredCellType <- gsub("/", ".", metadata$PredCellType)
colnames(countBySample) <- gsub("/", ".", colnames(countBySample))
countByMajor <- 
  sapply(unique(metadata$majorID), function(i){
    if(sum(metadata$majorID == i) > 1){
      return(rowSums(countBySample[, metadata$majorID == i]))
    }else{
      return(countBySample[, metadata$majorID == i])
    }
  })
# build DGEList for EdgeR

d = DGEList(counts = countBySample,
            samples = metadata,
            group = metadata$Tissue,
            remove.zeros = TRUE)

summary(d$samples$lib.size)
keep.samples <- d$samples$lib.size > 200000
table(keep.samples)

d <- d[, keep.samples]
keep.genes <- filterByExpr(d, group=d$samples$PredCellType)
table(keep.genes)
d <- d[keep.genes, , keep = FALSE]
#d <- calcNormFactors(d)
# plot MDS
cluster <- as.factor(d$samples$PredCellType)
plotMDS(d, pch=16, col = c(2:10)[cluster], main="MDS")
legend("topright", legend=levels(cluster),
pch=16, col=2:10, cex=0.8)

# perform DE tests across status one cell type at a time
tests <- list()
outputs <- list()
for (c in c("CD4_T", "CD8_T", "Memory_B", "Naive_B", "NK")) { #c("T", "B", "NK")) {
  sample_idx = which((d$samples$PredCellType == c) & (d$samples$Sample != "COB-8"))
  covariates = d$samples[sample_idx, ]
  y = d
  y$counts <- y$counts[, sample_idx]
  y$samples <- y$samples[sample_idx, ]
  Batch <- factor(covariates$batch) # only two batches
  Tissue <- factor(covariates$Tissue)
  design <- model.matrix(~ Batch + Tissue)
  rownames(design) <- colnames(y)
  y <- estimateDisp(y, design, robust = TRUE)
  fit <- glmQLFit(y, design, robust = TRUE)
  lrt <- glmQLFTest(fit)
  tests[[c]] <-
    as.data.frame(lrt$table)
  outputs[[c]] <- y
}
```

## Cross-tissue comparison

Different models for cross-tissue comparison:

* Y ~ group + batch
* Y ~ 0 + group + batch

Procedures:
1. Filtered out samples with total libray size < 20K (107/143 samples retained)
2. Filtered out peaks with low read counts (70% of reads retained)
3. Computed the scaling factor to normalize read depth by TMM
4. Estimated dispersion based on model design
5. Performed NB regression

```{r eval = F}
group <- factor(
  paste0(d$samples$major, ".",
         d$samples$Tissue))
colnames(d$counts) <- gsub("/", ".", colnames(d$counts))
y <- DGEList(d$counts, group=group)
keep <- filterByExpr(y)
summary(keep)
y <- y[keep, , keep.lib.sizes=FALSE]
y <- calcNormFactors(y)

plotMD(cpm(y, log=TRUE), column=3)
abline(h=0, col="red", lty=2, lwd=2)

plotMDS(y, pch=16, col = c(2:10)[group], main="MDS")
legend("topright", legend=levels(group),
pch=16, col=2:10, cex=0.8)

batch <- factor(d$samples$batch)
donor <- factor(d$samples$Donor)
design <- model.matrix(~ 1 + batch + group)
#colnames(design) <- gsub("donor", "", colnames(design))
#colnames(design) <- levels(group)
design
y <- estimateDisp(y, design, robust=TRUE)
plotBCV(y)
```

```{r eval = F}
### Replicate previous results
# TBD why including intercept makes the results so different?
y <- DGEList(d$counts[, d$samples$major == "B"], 
             group = d$samples$Tissue[d$samples$major == "B"])
keep <- filterByExpr(y)
summary(keep)
y <- y[keep, , keep.lib.sizes=FALSE]
y <- calcNormFactors(y)
batch <- factor(d$samples$batch[d$samples$major == "B"])
group <- factor(y$samples$group)
design <- model.matrix(~ batch + group)
y <- estimateDisp(y, design, robust=TRUE)
plotBCV(y)
```

```{r eval = F}
fit <- glmQLFit(y, design, robust=TRUE)
plotQLDisp(fit)
qlf <- glmQLFTest(fit)#, contrast=con)

cluster <- as.factor(d$samples$major)
qlf <- list()
for(i in 1:nlevels(cluster)){
  focalCT <- levels(cluster)[i]
  con <- makeContrasts(
    paste(glue::glue("group", focalCT, ".Lung"),
          glue::glue("group", focalCT, ".Spleen"), 
          sep = "-"),
    levels=design)
  qlf[[i]] <- glmQLFTest(fit, contrast=con)
}
```

```{r}
qlf <- readRDS("output/u19_output/diff_peak_analysis/cross_tissue_subsets_lymph_edgeR.RDS")
count_tbl <-
  sapply(qlf, function(i){
    topTags(i)
    plotMD(i)
    summary(decideTests(i))})
```

Notes:  
Memory B cells show more cross-tissue differences, while T subsets seem to violate model assumptions due to large variation

### Genomic region enrichment with rGREAT

```{r}
## A function to identify overlapped genes with selected GO terms
get_overlapping_sets <- function(res, GO_sets){
  library(org.Hs.eg.db)
  gr <- getRegionGeneAssociations(res)
  assoc_genes <- unlist(gr$annotated_genes)
  overlapped <- 
    lapply(GO_sets, function(i) {
        paste(unique(assoc_genes[names(assoc_genes) %in% res@gene_sets[[i]]]), 
              collapse = ",")
      })
  names(overlapped) <- GO_sets
  return(overlapped)
}
```


```{r eval = F}
# Genomic region enrichment with rGREAT

## default hg19
library(rGREAT)
set.seed(123)
gr <- proj@peakSet[as.numeric(rownames(qlf_lungs[[3]])[which(decideTests(qlf_lungs[[3]]) == -1)])]

res<- lapply(c("GO:BP", "GO:CC", "GO:MF", "msigdb:C7:IMMUNESIGDB",
                     "msigdb:C2:CP:KEGG", "msigdb:C5:HPO"), function(i) {
  return(great(gr, i, "hg38", background = proj@peakSet))
})

library(simplifyEnrichment)
sig_go_ids <- 
  res_list[[2]] %>% 
    filter(p_adjust < 0.001) %>%
      pull(id)
cl = simplifyGO(sig_go_ids)
```
**Clustering of GO terms**

![MemB GO BP terms](https://github.com/user-attachments/assets/c204a6dd-f905-4c63-a41a-412749cfe8dc)

![MemB GO MF terms](https://github.com/user-attachments/assets/886cdbf0-2c4a-477b-99c0-29c94aa1798d)


**A table of significant GO terms**

Differential peaks from lung memB cells are enriched for neuron fate commitment, T cell recombination and apoptotic processes

Differential peaks from lung memB cells are also enriched for immune receptor binding, metabolism, oxysterol binding, etc
```{r}
#res <- readRDS("output/u19_output/GO_enrichment/lung_control_specific_T_GO.RDS")
res <- readRDS("output/u19_output/GO_enrichment/lung_specific_memB_GO.RDS")
res_list <- lapply(res, function(i){getEnrichmentTable(i)})
res_genes <- lapply(res, function(i){getRegionGeneAssociations(i)})

titles <- c("GO: Biological Process", "GO: Molecular Functions")
for(idx in 1:2){
  
  selected <- 
    res_list[[idx]] %>% 
      filter(p_adjust < 0.05) %>%
        arrange(desc(fold_enrichment)) %>%
          pull(id)
  
  gene_set <- unlist(get_overlapping_sets(res[[idx]], selected[1:15]))

  print(
    DT::datatable(
      res_list[[idx]] %>% dplyr::select(id, description,
                    fold_enrichment, 
                    p_adjust, 
                    mean_tss_dist, 
                    gene_set_size) %>%
        filter(p_adjust < 0.05) %>%
        arrange(desc(fold_enrichment)) %>% slice_head(n = 15) %>%
        mutate(fold_enrichment = round(fold_enrichment, 1),
               p_adjust = format(p_adjust, scientific = TRUE, digit = 2),
               overlapped_genes = gene_set) %>%
        filter(duplicated(overlapped_genes) == FALSE),
        caption = titles[idx])
  )
}
```

### Cross-tissue comparison for asthma fine-mapped regions

```{r fig.height = 8, fig.width = 10}
qlf <- readRDS("output/u19_output/diff_peak_analysis/cross_tissue_subsets_lymph_edgeR.RDS")
gr_peaks<- readRDS("output/u19_output/u19_full_peaks_near_confSNPs.RDS")

gginput <- 
  sapply(qlf, function(i) {
    i$table[match(unique(gr_peaks$peakID[gr_peaks$type == "high"]),
                         rownames(i$table)), ]$PValue})
colnames(gginput) <- names(qlf)
rownames(gginput) <- 
  gr_peaks$labels[match(unique(gr_peaks$peakID[gr_peaks$type == "high"]),
                        gr_peaks$peakID)]
# remove NAs
gginput <- gginput[rowSums(is.na(gginput)) == 0, ]
# make a heatmap for DE results 
library(circlize)
col_fun = colorRamp2(c(0, 1, 2.5), c("blue", "white", "red"))
pval_heatmap <-
  Heatmap(-log10(gginput[, gsub("/", ".", ct_order)]), 
          row_split = gr_peaks$trait[match(rownames(gginput), gr_peaks$labels)],
          cluster_columns = FALSE,
          row_names_side = "left",
          show_row_dend = FALSE,
          name = "-log10(P-value)",
          na_col = "grey",
          #column_title = "control",
          row_title = NULL,
          #right_annotation = row_ha,
          width = nrow(gginput)*unit(2, "mm"),
          col = col_fun)

pval_heatmap
```
```{r}
pData <- 
  data.frame(gginput) %>% 
    dplyr::mutate(Peak = rownames(gginput)) %>%
      pivot_longer(!Peak, names_to = "CellType", values_to = "pval") %>%
        dplyr::filter(!CellType %in% c("Other", 
                              "Th17.CD4_T", 
                              "CD8.CD4_T"))

# perform multiple testing for all peaks
DT::datatable(
  pData %>%
    mutate(FDR = round(p.adjust(pval, method = "fdr"), 2),
           pval = format(pval, scientific = T, digits = 2)) %>%
    dplyr::filter(FDR < 0.1), 
  caption = "Selected peaks with FDR < 0.1")
```

## cross-status comparison for lung immune major lymphocytes

Lung T cells display hundreds of peaks with less chromatin accessibility in case than control.

```{r eval = F}
group <- factor(
  paste0(d$samples$PredCellType, ".",
         d$samples$Tissue, ".",
         d$samples$Status))

y <- DGEList(d$counts, group=group)

keep <- filterByExpr(y)
summary(keep)
y <- y[keep, , keep.lib.sizes=FALSE]
y <- calcNormFactors(y)

batch <- factor(d$samples$batch)
design <- model.matrix(~ 0 + group + batch)
#colnames(design) <- gsub("donor", "", colnames(design))
#colnames(design) <- levels(group)

y <- estimateDisp(y, design, robust=TRUE)
plotBCV(y)
fit <- glmQLFit(y, design, robust=TRUE)
plotQLDisp(fit)
```

```{r eval = F}
qlf_lungsub <- list()
cluster <- factor(d$samples$PredCellType)
for(i in 1:nlevels(cluster)){
  focalCT <- levels(cluster)[i]
  con <- makeContrasts(
    paste(glue::glue("group",focalCT, ".Lung.control"),
          glue::glue("group", focalCT, ".Lung.case"), 
          sep = "-"),
    levels=design)
  qlf_lungsub[[i]] <- glmQLFTest(fit, contrast=con)
}
saveRDS(qlf_lungsub, "../output/u19_output/diff_peak_analysis/cross_status_lung_subsets_lymph_edgeR.RDS")
qlf_spleens <- list()
cluster <- factor(d$samples$major)
for(i in 1:nlevels(cluster)){
  focalCT <- levels(cluster)[i]
  con <- makeContrasts(
    paste(glue::glue("group",focalCT, ".Spleen.control"),
          glue::glue("group", focalCT, ".Spleen.case"),
          sep = "-"),
    levels=design)
  qlf_spleens[[i]] <- glmQLFTest(fit, contrast=con)
}
saveRDS(qlf_spleens, "../output/u19_output/diff_peak_analysis/cross_status_spleen_major_lymph_edgeR.RDS")
```
```{r}
qlf_lungs <- readRDS("output/u19_output/diff_peak_analysis/cross_status_lung_major_lymph_edgeR.RDS")
count_tbl <-
  sapply(qlf_lungs, function(i){
    topTags(i)
    plotMD(i)
    summary(decideTests(i))})
```

### Genomic regions enrichment results

Differential peaks from control lung Tregs are enriched for multiple terms related to regulation of chemokine production.

```{r}
res <- readRDS("output/u19_output/GO_enrichment/lung_control_specific_T_GO.RDS")
res_list <- lapply(res, function(i){getEnrichmentTable(i)})
res_genes <- lapply(res, function(i){getRegionGeneAssociations(i)})

titles <- c("GO: Biological Process", 
            "GO: Cellular Component",
            "GO: Molecular Functions")
for(idx in c(1,3)){
  
  selected <- 
    res_list[[idx]] %>% 
      filter(p_adjust < 0.05) %>%
        arrange(desc(fold_enrichment)) %>%
          pull(id)
  
  gene_set <- unlist(get_overlapping_sets(res[[idx]], selected[1:length(selected)]))

  print(
    DT::datatable(
      res_list[[idx]] %>% dplyr::select(id, description,
                    fold_enrichment, 
                    p_adjust, 
                    mean_tss_dist, 
                    gene_set_size) %>%
        filter(p_adjust < 0.05) %>%
        arrange(desc(fold_enrichment)) %>% 
        mutate(fold_enrichment = round(fold_enrichment, 1),
               p_adjust = format(p_adjust, scientific = TRUE, digit = 2),
               overlapped_genes = gene_set) %>%
        filter(duplicated(overlapped_genes) == FALSE),
        caption = titles[idx])
  )
}
```

### Cross-status comparison for asthma fine-mapped regions

None of the peaks show significant differences in chromatin accessibility between case and control.

```{r fig.height = 8, fig.width = 10}
gginput <- 
  sapply(qlf_lungs, function(i) {
    i$table[match(unique(gr_peaks$peakID[gr_peaks$type == "high"]),
                         rownames(i$table)), ]$PValue})
#cluster <- factor(d$samples$major)
colnames(gginput) <- c("B", "NK", "T", "Other")
rownames(gginput) <- 
  gr_peaks$labels[match(unique(gr_peaks$peakID[gr_peaks$type == "high"]),
                        gr_peaks$peakID)]
# remove NAs
gginput <- gginput[rowSums(is.na(gginput)) == 0, ]
# make a heatmap for DE results 
library(circlize)
col_fun = colorRamp2(c(0, 1, 2.5), c("blue", "white", "red"))
pval_heatmap <-
  Heatmap(-log10(gginput),#[, gsub("/", ".", ct_order)]), 
          row_split = gr_peaks$trait[match(rownames(gginput), gr_peaks$labels)],
          cluster_columns = FALSE,
          row_names_side = "left",
          show_row_dend = FALSE,
          name = "-log10(P-value)",
          na_col = "grey",
          #column_title = "control",
          row_title = NULL,
          #right_annotation = row_ha,
          width = nrow(gginput)*unit(2, "mm"))
          #col = col_fun)

pval_heatmap
```

## Cell-type specific analyses

* Model: Y ~ group + batch + donor
* COB-8 sample and non-lymphocyte cells removed
* Almost half of peaks filtered due to relatively low read counts compared to others 
* Normalized by TMM

```{r eval = F}
d$samples$Donor <- 
  unlist(
    lapply(d$samples$Sample, function(i){
      sample_cov$ROBI[match(i, sample_cov$core_ID)]}))
# no Other group and COB-8 included in cell-type comparison
sample_idx <- which(d$samples$PredCellType != "Other" &
                 d$samples$Sample != "COB-8")
y <- DGEList(
  counts = d$counts[, sample_idx],
  samples = d$samples[sample_idx, ])

keep <- filterByExpr(y)
summary(keep)
y <- y[keep, , keep.lib.sizes=FALSE]
y <- calcNormFactors(y)

donor <- factor(y$samples$Donor)
cluster <- factor(y$samples$PredCellType)
batch <- factor(y$samples$batch)
design <- model.matrix(~ cluster + batch + donor)
colnames(design) <- gsub("donor", "", colnames(design))
colnames(design)[1] <- "Int"
ncls <- nlevels(factor(y$samples$PredCellType))
contr <- rbind( matrix(1/(1-ncls), ncls, ncls),
        matrix(0, ncol(design)-ncls, ncls) )
diag(contr) <- 1
contr[1,] <- 0
rownames(contr) <- colnames(design)
colnames(contr) <- paste0("cluster", levels(cluster))
contr
```

```{r eval = F}
y <- estimateDisp(y, design, robust=TRUE)
plotBCV(y)
fit <- glmQLFit(y, design, robust=TRUE)
plotQLDisp(fit)
qlf <- list()
for(i in 1:ncls){
  qlf[[i]] <- glmQLFTest(fit, contrast=contr[,i])
  qlf[[i]]$comparison <- paste0("cluster", levels(cluster)[i], "_vs_others")
}
```
```{r}
load("output/cross_celltypes_lymphOnly_edgeR.RData")
count_tbl <-
  sapply(new_qlf, function(i){
    topTags(i)
    plotMD(i)
    summary(decideTests(i))})
```

### Cell-type comparison for asthma fine-mapped regions

**EdgeR pseudobulk NB approch**


```{r eval = F}
lcpm <- cpm(y, log=TRUE)
annot <- data.frame(cluster=cluster)
rownames(annot) <- colnames(y)
ann_colors <- list(cluster=2:9)
names(ann_colors$cluster) <- levels(cluster)
row_idx <- gr_peaks$peakID[gr_peaks$type == "high"]
row_idx <- match(intersect(row_idx, rownames(lcpm)), rownames(lcpm))
top <- 10
topMarkers <- list()
for(i in 1:ncls) {
  ord <- order(qlf[[i]]$table$PValue, decreasing=FALSE)
  up <- qlf[[i]]$table$logFC[ord] > 0
  topMarkers[[i]] <- rownames(y)[ord[up][1:top]]
}
topMarkers <- unique(unlist(topMarkers))
pheatmap::pheatmap(lcpm[topMarkers, ], 
                   breaks=seq(-2,2,length.out=101),
  color=colorRampPalette(c("blue","white","red"))(100), scale="row",
  cluster_cols=TRUE, border_color="NA", fontsize_row=5,
  treeheight_row=70, treeheight_col=70, cutree_cols=7,
  clustering_method="ward.D2", show_colnames=FALSE,
  annotation_col=annot, annotation_colors=ann_colors)
```


Heatmap plot for comparison p-values
```{r fig.height = 8, fig.width = 10}

logfc_mat <- 
  sapply(new_qlf, function(i) {
    tbl = i$table[match(unique(gr_peaks$peakID[gr_peaks$type == "high"]),
                         rownames(i$table)), ]
    return(tbl$logFC/tbl$logCPM)})
pval_mat <- sapply(new_qlf, function(i) {
    i$table[match(unique(gr_peaks$peakID[gr_peaks$type == "high"]),
                         rownames(i$table)), ]$PVal})
#cluster <- factor(y$samples$PredCellType)
colnames(logfc_mat) <- names(new_qlf)
rownames(logfc_mat) <- 
  gr_peaks$labels[match(unique(gr_peaks$peakID[gr_peaks$type == "high"]),
                        gr_peaks$peakID)]
colnames(pval_mat) <- colnames(logfc_mat)
rownames(pval_mat) <- rownames(logfc_mat)

pval_mat <- pval_mat[rowSums(is.na(pval_mat)) == 0, ]
logfc_mat <- logfc_mat[rowSums(is.na(logfc_mat)) == 0, ]
# make a heatmap for DE results 
library(circlize)
#col_fun = colorRamp2(c(-0.5, 0, 0.5), c("blue", "white", "red"))
#mean_expr <- rowMeans(cpm(y, log = TRUE))
fc_heatmap <-
  Heatmap(#-log10([, gsub("/", ".", ct_order[ct_order != 'Other'])]), 
          logfc_mat[, gsub("/", ".", ct_order[ct_order != 'Other'])], 
          row_split = gr_peaks$trait[match(rownames(logfc_mat), gr_peaks$labels)],
          cluster_columns = FALSE,
          row_names_side = "left",
          show_row_dend = FALSE,
          #name = "-log10(P-value)",
          name = "logFC",
          na_col = "grey",
          #column_title = "control",
          row_title = NULL,
          #right_annotation = row_ha,
          width = nrow(logfc_mat)*unit(3, "mm"))
          #col = col_fun)


row_idx = 
  gr_peaks$peakID[match(rownames(logfc_mat)[unlist(row_order(fc_heatmap))], gr_peaks$labels)]
row_ha = 
  rowAnnotation(AvgCPMs = anno_barplot(mean_expr[match(row_idx, names(mean_expr))]))
col_fun = colorRamp2(c(0, 2, 10), c("blue", "white", "red"))
pval_heatmap <-
  Heatmap(-log10(pval_mat[, gsub("/", ".", ct_order[ct_order != 'Other'])]), 
          #logfc_mat[, gsub("/", ".", ct_order[ct_order != 'Other'])], 
          row_split = gr_peaks$trait[match(rownames(pval_mat), gr_peaks$labels)],
          cluster_columns = FALSE,
          #row_order = unlist(row_order(fc_heatmap)),
          row_names_side = "left",
          show_row_dend = FALSE,
          name = "-log10(P-value)",
          #name = "logFC",
          na_col = "grey",
          #column_title = "control",
          row_title = NULL,
          right_annotation = row_ha,
          width = nrow(pval_mat)*unit(3, "mm"),
          col = col_fun)

```

```{r, fig.height = 8, fig.width = 12}
fc_heatmap + pval_heatmap
```



