---
title: "Linking asthma fine-mapped regions with lung immune OCRs"
output: html_document
date: '`r Sys.Date()`'
---

# Objective

For asthma fine-mapped regions, what cell types have open chromatin at these regions?

* Shared or cell-type specific?
* Present In lungs or spleens?

For the selected regions, what genes are differentially expressed for cell types with open chromatin?

# Procedure

What we have: 

  * asthma fine-mapped variants grouped by enhancers using multiple functional data
  * normalized CA and GE matrix (Feature x cell)
    - TMM and log-normalization (column-wise)
    - scaled to zscore within each tissue (row-wise)

1. Define asthma candidate enhancers by ePIPs $\ge$ 0.5
  * the selected cutoff shows strong enrichment with overlapping OCRs
  * randomly sampling 10 enhancers from low ePIPs as negative ctrls 
2. Overlap with the joint set of peaks in our dataset
  * peaks need to contain fine-mapped variants
3. Compute average levels across cell types 
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)

library(ArchR)
library(dplyr)
library(tidyr)
library(rtracklayer)
library(GenomicRanges)
library(scales)
library(edgeR)
library(BSgenome.Hsapiens.UCSC.hg38)
library(ComplexHeatmap)
#library(plotly)
#source("~/projects/lung_immune_fine_mapping/code/archR_helper.r")
set.seed(1)
options(repr.plot.width=11,.plot.height=8)

addArchRGenome("hg38")
work.dir<-"~/projects/u19_multiomics/data/cellranger_outs"

proj <- loadArchRProject("~/cluster/projects/u19_multiomics/analyses/u19_peaks_tissue/")
ct_order <-
  c(
  "Naive_B",
  "Memory_B",
  "NK",
  "CD8_T",
  "CD8/CD4_T",
  "CD4_T",
  "Th17/CD4_T",
  "Treg",
  "Other"
)

```

```{r eval = F}
# group count matrix by pseudo-bulk

peakMatrix$CellGroup2 <-
  paste(peakMatrix$Sample,
        peakMatrix$PredCellType,
        sep = "_")

countBySample <- 
  sapply(unique(peakMatrix$CellGroup2), function(i) {
  rowSums(assays(peakMatrix[, peakMatrix$CellGroup2 == i])$PeakMatrix)
})

countBySample <- 
  sapply(unique(peakMatrix$CellGroup2), function(i) {
  colSums(binary[peakMatrix$CellGroup2 == i, ])/sum(peakMatrix$CellGroup2 == i)
})

metadata <-
  data.frame(colData(peakMatrix), row.names = NULL) %>%
  select(Sample, PredCellType, Tissue, Batch, Sex, Age, Race, Status) %>% distinct()
metadata$Status <-
  factor(metadata$Status, levels = c("control", "case"))
```

```{r}
normalize_counts <- function(input){
  # build a DGEList with EdgeR
  d <- DGEList(counts = input, 
               remove.zeros = TRUE)
  # compute effective library size through TMM
  d <- calcNormFactors(d)
  
  # log-normalization
  counts <- cpm(d, normalized.lib.sizes = TRUE, log = TRUE)
  return(counts)
}

aggregate_counts <- function(counts, aggr_by){
   aggr <-
     sapply(levels(aggr_by), function(j) {
        sub_counts <-
          counts[, grepl(j, colnames(counts))] 
        return(rowMeans(sub_counts))
    })
   return(aggr)
}
```

```{r}
f <- fread("~/projects/lung_lymph_scMultiomics/data/CRE_0_asthma_fine-mapping_summary.txt")
# add SNP positions 
# ukbb_snps <- fread("/home/jinggu/cluster/data/gwas/ukbb_vars/ukb_b38_vars.txt.gz")
# f$pos <-
#   unlist(lapply(f$snps, function(i){
#     snps <- strsplit(i, ",")[[1]]
#     positions <- paste(ukbb_snps[ukbb_snps$id %in% snps, ]$pos, collapse = ",")
#     return(positions)
#   }))
# f[f$pos == ""] <- 137605991 # fill in the empty position
# write.table(f, "../data/CRE_0_asthma_fine-mapping_summary.txt", quote = F, sep = "\t", row.names = F)

```


```{r}
# Overlap asthma fine-mapped regions with OCRs in our dataset
gr <- 
  GRanges(seqnames = paste0("chr", f$chr), 
            ranges = IRanges(start = f$start,
                             end = f$end),
            aoa_epip = f$aoa_epip,
            coa_epip = f$coa_epip,
            pos = f$pos
           )

# liftover to hg38
ch=import.chain("~/resources/genomes/chain_files/hg19ToHg38.over.chain")

seqlevelsStyle(gr) <- "UCSC"
gr_hg38 <- unlist(liftOver(gr, ch))
gr_hg38$labels <- paste(round(gr_hg38$aoa_epip, 2),
                        round(gr_hg38$coa_epip, 2),
                         sep = "_")

gr_hg38$type <- ifelse(gr_hg38$aoa_epip >= 0.5|gr_hg38$coa_epip >= 0.5, "high", "low")

paired <- findOverlaps(proj@peakSet, gr_hg38)

# calculate the distances btw fine-mapped SNPs and the overlapped peak sets
gr_peaks <- proj@peakSet[paired@from]
gr_peaks$distToSNP <- unlist(
    lapply(1:length(paired), function(i){
    snps <- strsplit(gr_hg38[paired@to[i]]$pos, ",")[[1]]
    gr_snps <- 
      GRanges(seqnames = 
                rep(as.character(gr_hg38[paired@to[i]]@seqnames), length(snps)),
              IRanges(start = snps))
    dist_gr = distanceToNearest(proj@peakSet[paired@from[i]], gr_snps)
    if(length(dist_gr) < 1){
      return(NA)
    }else{
      return(dist_gr@elementMetadata$distance)
    }
  }))

# filter out peaks that do not contain fine-mapped SNPs
paired <- paired[gr_peaks$distToSNP == 0]
gr_peaks <- gr_peaks[gr_peaks$distToSNP == 0]
gr_peaks$type <- gr_hg38[paired@to]$type

```
# Result

## Enrichment test for enhancers with high ePIPs

At a cutoff of 0.5 for ePIPs, enhancers with high ePIPs show significant enrichment for the overlapped OCRs. 
```{r}
fisher_x = rbind(table(gr_hg38[unique(paired@to)]$type), 
                 table(gr_hg38[-paired@to, ]$type))
colnames(fisher_x) <- c("high_ePIP", "low_ePIP")
rownames(fisher_x) <- c("overlapped", "not_overlapped")
fisher_x
fisher.test(fisher_x)
```

## Size distribution for Ethan's list of enhancers

* 27 enhancers with PIPs >= 0.5
* 31 peaks (501 bp) overlap with 27 enhancers at high PIPs
* The above peaks are required to contain asthma fine-mapped variants.

```{r}
gr_hg38_paired <- gr_hg38[paired@to]
hist(data.frame(ranges(gr_hg38_paired[gr_hg38_paired$type == "high"]))$width,
     main = "Size distribution for enhancers with high ePIPs",
     xlab = "peak size")
```

## Processing chromatin accessibility separately for each tissue

We computed z-scores separately for each tissue.
```{r}
# obtain count data with selected peaks 
load("output/u19_output/u19_full_atac_by_tissue-celltype_pseudobulk.RData")
metadata$PredCellType <-
  factor(metadata$PredCellType,
         levels = ct_order)

counts <- normalize_counts(countBySample)

# compute z-scores separately for each tissue
lung_counts<- counts[, metadata$Status == "control" & metadata$Tissue == "Lung"]
lung_counts <- t(scale(t(lung_counts)))

spleen_counts <- counts[, metadata$Status == "control" & metadata$Tissue == "Spleen"]
spleen_counts <- t(scale(t(spleen_counts)))

aggr_lung <- aggregate_counts(lung_counts[paired@from, ], metadata$PredCellType)
aggr_spleen <- aggregate_counts(spleen_counts[paired@from, ], metadata$PredCellType)
```

### Comparing chromatin accessibility between peaks at high or low ePIPs 

 * Lung CD4_T subsets show overall higher mean accessibility btw peaks with high ePIP and those with low ePIPs
    - Treg has the highest mean difference 
 * Spleen subsets show not much differences between peaks with high ePIPs and those with low ePIPs.
 
```{r, fig.width = 9}
aggrDF <- rbind(
  data.frame(feature = "lungs", aggr_lung),
  data.frame(feature = "spleens", aggr_spleen))
  
gginput <- 
  data.frame(aggrDF, 
             type = rep(gr_hg38[paired@to]$type, 2)) %>%
  pivot_longer(!c(type, feature), 
               values_to = "counts",
               names_to = "CellType") %>%
  mutate(CellType = factor(CellType,
                           levels = gsub("/", ".", ct_order)))

ggplot(gginput,
       aes(y = counts,
           x = type,
           fill = type)) + 
  geom_boxplot() + 
  theme_bw() + xlab("") + ylab("normalized accessibility") + 
  labs(fill = "ePIP level") + 
  facet_grid( feature ~ CellType)

```

```{r}
# function for making heatmap plot
plot_heatmap <- function(data){
  # color_lim =
  #   min(unlist(lapply(split(data, data$feature), function(i) {
  #     max(i$counts)
  #   })))
  p <- ggplot(data,
            aes(x = cell_type, 
                y = labels,
                fill = counts)) +
  geom_tile() + xlab("") +
  labs(title = "Chromatin accessibility landscape nearby enhancers with high ePIPs ",
       y = "enhancer-gene pairs",
       fill = "chromatin accessibility") +
  # scale_fill_viridis_b(
  #   limits = c(0, color_lim),
  #   breaks = round(as.numeric(quantile(data$counts)), 1)
  #   ) +
  scale_fill_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0) + 
  #scale_fill_viridis_c() +
  # scale_fill_viridis_b() + 
  theme(
    axis.text.x = element_text(
      size = 12,
      angle = 45,
      vjust = 0.6
    ),
    axis.text.y = element_text(size = 12),
    strip.text.x = element_text(size = 12)
  )  + facet_grid( ~ feature)#, scales = "free")
  return(p)
}
```

```{r }
# retain one peak per candidate enhancer with maximum level
labels <-
  paste(as.character(gr_peaks),
                     gr_hg38[paired@to]$labels, 
                     gr_peaks$nearestGene,
                     sep="_")
gginput <-
  data.frame(
    aggrDF, 
    labels = rep(labels, 2),
    type = rep(gr_hg38[paired@to]$type, 2),
    row.names = NULL) %>% 
      rowwise() %>%
      mutate(mean_value = mean(c_across(where(is.numeric)))) %>%
      group_by(labels, feature) %>%
      dplyr::filter(mean_value == max(mean_value))


gginput <-
  gginput %>% ungroup() %>%
  select(!mean_value) %>%
  pivot_longer(
    cols = !c(labels, type, feature), #, feature),
    names_to = "cell_type",
    values_to = "counts"
  )
```

## Examining chromatin accessibility for asthma fine-mapped regions 

```{r fig.width = 12, fig.height = 8}
gginput_avg <-
  gginput %>% group_by(feature, cell_type) %>% 
    dplyr::summarise(counts = mean(counts)) %>%
    mutate(labels = "average_accessibility",
           type = "average") %>%
    select(feature, labels, type, cell_type, counts)

mat <-
  data.frame(rbind(gginput, gginput_avg)) %>% 
    pivot_wider(names_from = cell_type, 
                values_from = counts) %>%
    mutate(type = factor(type, levels = c("low", "high", "average")))

# randomly sample 10 peaks in low ePIPs
set.seed(10)
idx <- sample(1:sum(mat$feature == "lungs" & mat$type == "low"), 
              10, replace = FALSE)
lung_mat <- mat[mat$feature == "lungs", ]
lung_mat <- 
  rbind(lung_mat[lung_mat$type != "low", ],
        lung_mat[lung_mat$type == "low", ][idx, ])

types <- lung_mat$type
labels <- lung_mat$labels
lung_mat<-as.matrix(lung_mat[, c(-3:-1)])
rownames(lung_mat) <- labels

lung_heatmap<-
  Heatmap(lung_mat, row_split = types,
        cluster_columns = FALSE,
        row_names_side = "left",
        show_row_dend = FALSE,
        name = "lungs",
        column_title = "lungs",
        row_title = NULL,
        width = nrow(lung_mat)*unit(2, "mm"))


spleen_mat <- mat[mat$feature == "spleens", ]
spleen_mat <-  
  rbind(spleen_mat[spleen_mat$type != "low", ],
        spleen_mat[spleen_mat$type == "low", ][idx, ])

types <- spleen_mat$type
labels <- spleen_mat$labels
spleen_mat<-as.matrix(spleen_mat[, c(-3:-1)])
rownames(spleen_mat) <- labels

spleen_heatmap <-
  Heatmap(spleen_mat, row_split = types,
          cluster_columns = FALSE,
          row_dend_side = "right",
          row_names_side = "left",
          name = "spleens",
          column_title = "spleens",
          row_title = NULL,
          width = nrow(spleen_mat)*unit(2, "mm"))

```

### Heatmaps for accessibiltiy across cell types in normal tissues

* Top section: 10 randomly sampled peaks overlapping enhancers with low ePIPs
* Middle section: average accessibility for all peaks overlapped regardless of PIPs
* Bottom section: all peaks overlapping enhancers with high ePIPs

Y-axis: peak coordinate in hg38, aoa-ePIP, coa-ePIP, nearest gene
```{r fig.width = 14, fig.height = 10}
lung_heatmap + spleen_heatmap
```
* Overall, we see more chromatin near asthma fine-mapped variants are more open in lungs than spleens.
* shared across all lymphocytes
  - IL2RA, SMAD3, BACH2, RPS26
* shared across T subsets
  - GAL3ST2, CXCR5, CCR7, 
* specific to NK/CD8 T cells:
  - IRF1, TM9SF2, TSLP, LINC00298, TBX21,
* specific to Treg:
  - TNFRSF11A, SLC7A10 (both specific to lungs)


### Heatmaps for accessibiltiy across cell types in case and control

```{r}
# compute z-scores separately for case and control
case_counts <- counts[, metadata$Status == "case" & metadata$Tissue == "Lung"]
case_counts <- t(scale(t(case_counts)))

aggr_case <- aggregate_counts(case_counts[paired@from, ], metadata$PredCellType)

```

```{r}
# retain one peak per candidate enhancer with maximum level
gginput <- 
  data.frame(
    aggr_case,
    labels = paste(as.character(gr_peaks),
                   gr_hg38[paired@to]$labels, 
                   gr_peaks$nearestGene,
                   sep="_"),
    type = gr_hg38[paired@to]$type,
    row.names = NULL) %>%
    rowwise() %>%
    mutate(mean_value = mean(c_across(where(is.numeric)))) %>%
    group_by(labels) %>%
    dplyr::filter(mean_value == max(mean_value))

gginput <-
  gginput %>% ungroup() %>%
  select(!mean_value) %>%
  pivot_longer(
    cols = !c(labels, type), 
    names_to = "cell_type",
    values_to = "counts"
  ) 
gginput <- rbind(gginput, 
                 gginput_avg[gginput_avg$feature == "lungs", -1])
```

```{r, fig.width = 12, fig.height = 8}
mat <-
  data.frame(gginput) %>%
    pivot_wider(names_from = cell_type, 
                values_from = counts) %>%
    mutate(type = factor(type, levels = c("low", "high", "average")))

case_mat <- mat[match(rownames(lung_mat), mat$labels), ]
types <- case_mat$type
labels <- case_mat$labels
case_mat<-as.matrix(case_mat[, c(-2:-1)])
rownames(case_mat) <- labels

lung_heatmap<-
  Heatmap(lung_mat, row_split = types,
        cluster_columns = FALSE,
        row_names_side = "left",
        show_row_dend = FALSE,
        name = "lungs",
        column_title = "lungs_ctrl",
        row_title = NULL,
        width = nrow(lung_mat)*unit(2, "mm"))

case_heatmap<-
  Heatmap(case_mat, 
          cluster_columns = FALSE,
          row_dend_side = "right",
          row_names_side = "left",
          show_row_dend = FALSE,
          name = "case",
          column_title = "lung_case",
          width = nrow(case_mat)*unit(2, "mm"))
```

```{r fig.width = 14, fig.height = 10}
lung_heatmap + case_heatmap
```
Overall, chromatin levels in control are more open than those in case. Majoriy of CD4 subsets have lower chromatin accessibility, and NK/CD8+ T cells show higher accessibility near several regions, which turns out to be open in negative controls as well.


## Validate cell-type specific OCRs near asthma risk enhancers

Do their nearby genes show differential expression?
```{r}
rna_counts <- readRDS("output/fastTopics/aggregated_lymph_scRNA_counts.RDS")
rna_metadata <- readRDS("output/u19_output/u19_rna/u19_rna_cell_metadata.RDS")

rna_metadata$CellGroup <- paste(rna_metadata$core_ID, rna_metadata$major, sep="_")
# group count matrix by pseudo-bulk
countBySample <- 
  sapply(unique(rna_metadata$CellGroup), function(i) {
    if(sum(rna_metadata$CellGroup == i) < 2){
      return(rna_counts[rna_metadata$CellGroup == i, ])
    }else{
      return(colSums(rna_counts[rna_metadata$CellGroup == i, ]))
    }
})

metadata <- left_join(
  data.frame(CellGroup = colnames(countBySample)),
  rna_metadata %>% mutate(Sample = core_ID) %>%
    select(CellGroup, batch, major, Sample, tissue.ident) %>% 
    distinct(),
  by = "CellGroup")

d <- DGEList(counts=countBySample[, metadata$tissue.ident == "lungs" &
                                    metadata$Sample != "SMO-8"], 
             remove.zeros = TRUE)

d <- calcNormFactors(d) # for differential analysis
```

```{r}
rna_counts <- cpm(d, normalized.lib.sizes = TRUE, log = TRUE)#FALSE)
rna_counts <- t(scale(t(rna_counts)))
row_idx <- match(gr_peaks[gr_peaks$type == "high"]$nearestGene, rownames(rna_counts))
row_idx <- unique(row_idx[!is.na(row_idx)])
rna_counts <- 
  data.frame(
    aggregate_counts(
      counts = rna_counts[row_idx, ], 
      aggr_by = metadata$major))
```

### Overlapping DE genes across cell types with the nearest genes from cell-type specific OCRs

Cell-type specific OCRs are defined as those have higher accessibility in CD4 T subsets.

```{r}
fpath <-"~/projects/lung_immune_fine_mapping/output/DE_analysis_mast_minpct_0.1/marker_genes_batch_corrected.csv"
f <- fread(fpath) %>%
  filter(p_val_adj < 0.05 & avg_log2FC > 0.58)
colnames(f)[1] <- c("Gene")
sharedOCR_genes <- c("SMAD3", "IL2RA", "RPS26", "HDAC7", "BACH2", "TNFSF4", 
                     "OVOL1-AS1", "FASLG", "TSLP", "TLR10", "CCR7")
specific_genes <- setdiff(gr_peaks[gr_peaks$type == "high"]$nearestGene,
                          sharedOCR_genes)
DT::datatable(
  f %>% 
    filter(Gene %in% specific_genes) %>%
    mutate(avg_log2FC = round(avg_log2FC, 2),
           p_val_adj = format(p_val_adj, scientific = T, digits = 2),
           mean_cluster_accessibility = c("high", "high", "low", "low", "high", "high") ) %>%
    select(-c(p_val)), 
  caption = "Cell-type marker genes with cell-type specific open chromains")
```
### Heatmap for the expression of cell-type specific genes 

The scaling of gene expression was performed separately for each tissue, which agrees with the differential GE analyses done for lungs only. As a result, we see consistency in the differential expression and the heatmap of normalized expression. 
```{r fig.height = 8}
### make a heatmap plot

GE_heatmap<-
  Heatmap(rna_counts, 
          cluster_columns = FALSE,
          row_dend_side = "right",
          row_names_side = "left",
          show_row_dend = TRUE,
          name = "Gene Expression",
          column_title = "Gene expression",
          width = nrow(rna_counts)*unit(3, "mm"))
GE_heatmap
```


```{r eval = F}
library(edgeR)

Batch <- factor(metadata$batch)
Sample <- factor(metadata$Sample)
CellType <- factor(metadata$major)

design <- model.matrix( ~ Batch  + CellType)
rownames(design) <- colnames(d)
design

d <- estimateDisp(d, design, robust=TRUE)
fit <- glmFit(d, design)
lrt <- glmLRT(fit)

de <- decideTestsDGE(lrt)
summary(de)

plotMD(lrt)
abline(h=c(-1, 1), col="blue")
result_paired <- as.data.frame(topTags(lrt, n = nrow(counts)))
```


